The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-23 conf_article/neon.tex
Added Schatten, revised preliminaries

%------------------------------------------------------------------------------------
% \epsilon is predefined; override to use \varepsilon
\renewcommand{\epsilon}{\varepsilon}
\newcommand{\Rmn}{\R^{m\times n}}
\newcommand{\cB}{\mathcal{B}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}}
\newcommand{\cS}{\mathcal{S}}
\newcommand{\Ed}[2]{\mathbb{E}_{#1}\left[#2\right]}
\usepackage{mathtools}
\DeclarePairedDelimiter{\sqne}{\|}{\|_2^2}
%\DeclarePairedDelimiter{\norme}{\|}{\|_2}
\DeclarePairedDelimiter{\normf}{\|}{\|_\mathrm{F}}
\DeclarePairedDelimiter{\normkfk}{\|}{\|_\mathrm{KF-k}}
\DeclarePairedDelimiter{\normfstar}{\|}{\|_\mathrm{F*}}
\DeclarePairedDelimiter{\normftwo}{\|}{\|_\mathrm{F2}}
\DeclarePairedDelimiter{\sqns}{\|}{\|_{\mathrm{op}}^2}
\DeclarePairedDelimiter{\norms}{\|}{\|_{\mathrm{op}}}
\DeclarePairedDelimiter{\sqnn}{\|}{\|_{\mathrm{nuc}}^2}
\DeclarePairedDelimiter{\sqnf}{\|}{\|_{\mathrm{F}}^2}
\DeclarePairedDelimiter{\normn}{\|}{\|_{\mathrm{nuc}}}
\def\<#1,#2>{\langle #1,#2\rangle}
\DeclarePairedDelimiter{\dotprod}{\langle}{\rangle}

@@

Training a neural network is essentially an optimization of a function of several weight matrices. Let us start by considering the problem of minimizing a differentiable function of a {\it single} matrix:
\begin{equation}\label{eq:matrix_problem}
	F(\cdot)\colon \Rmn \to \R\,,\qquad F(\mX) \to \min_{\mX \in \Rmn}\,
\end{equation}
We equip the matrix space $\Rmn$ with a standard dot product $\<\cdot, \cdot> \to \R$ and a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which does not have to coincide with the Frobenius norm $\normf{\cdot}$. The dual norm $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ associated with $\norm{\cdot}$ is defined as
\begin{equation}\label{eq:dual_norm}
	\norm{\mG}^\dagger = \sup_{\mD \in \Rmn\,:\norm{\mD}\leq 1} \<\mG,\mD>\,.
\end{equation}

Such problems can be solved with an iterative algorithm based on the Linear Minimization Oracle (LMO):
\begin{equation}\label{eq:lmo}
	\mathrm{LMO}(\mG) \in \argmin_{\mD \in \cS} \<\mG, \mD>\,,
\end{equation}
where $\mG$ is typically a gradient (possibly with momentum) of $F$ and $\cS \subset \Rmn$ is some set. The simple update formula then can look like
\begin{equation}\label{eq:simple_update}
	\mX^{k + 1} = \mX^k + \gamma_k\mathrm{LMO}\left(\nabla F\left(\mX^k\right)\right)\,.
\end{equation}

Here we are particularly interested in the case when $\cS$ is a ball in the $\norm{\cdot}$ norm:
\begin{equation}\label{eq:ball}
	\cS = \cB_\eta = \{ \mD \in \Rmn \mid \norm{\mD} \leq \eta \}\,.
\end{equation}
In this case it can be easily shown that
\begin{equation} \label{eq:lmo_in_norm_ball}
	\argmin_{\mD \in \cS} \<\mG, \mD> = - \eta \{ \mD \in \cB_1 \mid \<\mG, \mD> = \norm{\mG}^\dagger\}\,.
\end{equation}
And update for $\mX$ in \cref{eq:usgd_update} becomes
\begin{equation}\label{eq:our_update}
	\mX^{k+1} = \mX^{k} - \gamma_k\eta \{\mD \in \cB_1 \mid \<\mM^k, \mD> = \norm{\mM^k}^\dagger\}\,.
\end{equation}
This formula will later allow us to easily compute updates for various norms.

In the next subsection we will derive and discuss updates induced by the choice of the norm in \cref{eq:our_update}. To describe those updates, the singular value decomposition (SVD) of $\mM^k$ is required. We denote SVD of $\mM^k$ as $\mM^k = \mU \mSigma \mV^\top$, where $\mU = [u_1, u_2, \dots, u_r]$, $\mSigma = \diag(\sigma_1, \sigma_2, \dots, \sigma_r)$, and $\mV = [v_1, v_2, \dots, v_r]$.

When $\norm{\cdot} = \norms{\cdot}$, \cref{eq:our_update} turns into:

\begin{equation}\label{eq:muon_update}
	\mX^{k+1} = \mX^{k} - \eta U V^\top
\end{equation}.

And when $\norm{\cdot} = \normf{\cdot}$, \cref{eq:our_update} turns into:

	\begin{equation}\label{eq:nsgd_update}
		\mX^{k+1} = \mX^{k} - \eta \frac{\mM^k}{\normf{\mM^k}}
	\end{equation}

\section{Duals to Ky Fan norms instead of the spectral norm}
%\subsection{Non-operator Norms $\norm{\cdot}$ lead to new updates}

\subsection{\texorpdfstring{$\normn{\mM^{k}}$ and Neon}{Neon}}
\begin{lemma}\label{lemma:neon_update}
	When $\norm{\cdot} = \normn{\cdot}$, \cref{eq:our_update} turns into:

	\begin{equation}\label{eq:neon_update}
		\mX^{k+1} = \mX^{k} - \eta u_1 v_1^\top
	\end{equation}
\end{lemma}
We name the derived algorithm \emph{Neon}. In the section ~\cref{sec:matrix-side}, we will discuss how to compute the update efficiently.

@@

Now: what will we get if we try to generalize?

We remind the reader that the Ky Fan k-norm (\citet{bhatia2013matrix}, p. 92), which we denote as $\normkfk{\cdot}$, is the sum of the k largest singular values of the matrix. It can be proved that $\normkfk{\cdot}^\dagger = \max\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\}$ (see \citet{bhatia2013matrix}, p. 96). Special cases of the Ky Fan k-norm are the Ky Fan 1-norm, which is the sprectral norm, and the Ky Fan $\min\{m, n\}$-norm, which is the nuclear norm.

@@

@@

\subsection{Add SignSGD: S-Muon and S-Neon}
Trivial formulas, but a need for more experiments.

@@

Proof of \cref{lemma:neon_update}:
Since $\norm{\cdot}^\dagger = \norms{\cdot}$, the goal is to reach $\norms{\mM^k} = \sigma_1$.

Let us note that $\Delta = u_1 v_1^\top$ delivers this value. Indeed, by the trace property and singular vectors orthogonality, $\<\mM^k, \Delta> =\<\mU \mSigma \mV^\top, u_1 v_1^\top> = \tr \diag({\sigma_1, 0, \dots, 0}) = \norms{\mM^k}$, which completes the proof.

>>>>

%------------------------------------------------------------------------------------
% \epsilon is predefined; override to use \varepsilon
\renewcommand{\epsilon}{\varepsilon}
\newcommand{\Rmn}{\R^{m\times n}}
\newcommand{\cB}{\mathcal{B}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}}
\newcommand{\cS}{\mathcal{S}}
\newcommand{\Ed}[2]{\mathbb{E}_{#1}\left[#2\right]}
\usepackage{mathtools}
\DeclarePairedDelimiter{\sqne}{\|}{\|_2^2}
%\DeclarePairedDelimiter{\norme}{\|}{\|_2}
\DeclarePairedDelimiter{\normf}{\|}{\|_\mathrm{F}}
\DeclarePairedDelimiter{\normkfk}{\|}{\|_{\text{KF-}k}}
\DeclarePairedDelimiter{\normfstar}{\|}{\|_\mathrm{F*}}
\DeclarePairedDelimiter{\normftwo}{\|}{\|_\mathrm{F2}}
\DeclarePairedDelimiter{\sqns}{\|}{\|_{\mathrm{2}}^2}
\DeclarePairedDelimiter{\norms}{\|}{\|_{\mathrm{2}}}
\DeclarePairedDelimiter{\sqnn}{\|}{\|_{\mathrm{nuc}}^2}
\DeclarePairedDelimiter{\sqnf}{\|}{\|_{\mathrm{F}}^2}
\DeclarePairedDelimiter{\normn}{\|}{\|_{\mathrm{nuc}}}
\def\<#1,#2>{\langle #1,#2\rangle}
\DeclarePairedDelimiter{\dotprod}{\langle}{\rangle}

@@

Training a neural network is essentially an optimization of a function of several weight matrices and a few vectors. Let us start by considering the problem of minimizing a differentiable function of a {\it single} matrix:
\begin{equation}\label{eq:matrix_problem}
	F(\cdot)\colon \Rmn \to \R\,,\qquad F(\mX) \to \min_{\mX \in \Rmn}\,
\end{equation}
We equip the matrix space $\Rmn$ with a standard dot product $\<\cdot, \cdot> \to \R$ and a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which does not have to coincide with the Frobenius norm $\normf{\cdot}$. The dual norm $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ that is associated with $\norm{\cdot}$ is defined as
\begin{equation}\label{eq:dual_norm}
	\norm{\mG}^\dagger = \sup_{\mD \in \Rmn\,:\norm{\mD}\leq 1} \<\mG,\mD>\,.
\end{equation}

Such problems can be solved with an iterative algorithm based on the Linear Minimization Oracle (LMO):
\begin{equation}\label{eq:lmo}
	\mathrm{LMO}(\mM^k) \in \argmin_{\mD \in \cS} \<\mM^k, \mD>\,,
\end{equation}
where $\mM^k$ is a gradient (or a momentum) of $F$ in $\mX^k$ and $\cS \subset \Rmn$ is some set. The update of the algorithm is defined as follows:
\begin{equation}\label{eq:simple_update}
	\mX^{k + 1} = \mX^k + \gamma_k\mathrm{LMO}\left(\mM\right)\,.
\end{equation}

We are particularly interested in the case when $\cS$ is a ball in the norm $\norm{\cdot}$:

$$\cS = \cB_\eta \in \{ \mD \in \Rmn \mid \norm{\mD} \leq \eta \}.$$
In this case,

$$\argmin_{\mD \in \cS} \<\mM^k, \mD> = - \eta \{ \mD \in \cB_1 \mid \<\mM^k, \mD> = \norm{\mM^k}^\dagger\}$$

and update for $\mX^k$ in \cref{eq:simple_update} simplifies to
\begin{equation}\label{eq:our_update}
	\mX^{k+1} = \mX^{k} - \eta_k \{\mD \in \cB_1 \mid \<\mM^k, \mD> = \norm{\mM^k}^\dagger\}\,
\end{equation}

By this formula, it is easy to compute updates for algorithms induced by various norms $\norm{\cdot}$. For example, when the norm $\norm{\cdot}$ is the Frobenius norm $\normf{\cdot}$, \cref{eq:our_update} turns into

\begin{equation}\label{eq:nsgd_update}
	\mX^{k+1} = \mX^{k} - \eta_k \frac{\mM^k}{\normf{\mM^k}},
\end{equation} which recovers Normalized SGD (NSGD). When one has the spectral norm $\norms{\cdot}$, one gets

\begin{equation}\label{eq:muon_update}
	\mX^{k+1} = \mX^{k} - \eta_k U V^\top,
\end{equation} which is Muon without the $\sqrt{m/n}$ factor.
Here, $\mM^k = \mU \mSigma \mV^\top$ is the Singular Value Decomposition (SVD) of $\mM^k$ ($\mU = [u_1, u_2, \dots, u_r]$, $\mSigma = \diag(\sigma_1, \space \sigma_2, \dots, \sigma_r)$, and $\mV = [v_1, v_2, \dots, v_r]$). Muon can be recovered by taking the RMS-to-RMS operator norm $\sqrt{\frac{n}{m}}\norms{\cdot}$.

\section{Duals to Ky Fan norms instead of the spectral norm}

\subsection{\texorpdfstring{$\normn{\mM^{k}}$ and Neon}{Neon}}
After having considered $\normf{\mM^k}$ and $\norms{\mM^k}$, it is natural to look at the nuclear norm $\normn{\mM^k}$.
\begin{lemma}\label{lemma:neon_update}
	When $\norm{\cdot} = \normn{\cdot}$, \cref{eq:our_update} turns into:

	\begin{equation}\label{eq:neon_update}
		\mX^{k+1} = \mX^{k} - \eta_k u_1 v_1^\top
	\end{equation}
\end{lemma}
\begin{proof}
Since $\norm{\cdot}^\dagger = \norms{\cdot}$, the goal is to reach $\norms{\mM^k} = \sigma_1$.

Let us note that $\Delta = u_1 v_1^\top$ delivers this value. Indeed, $\normn{\Delta}=1$ and by the trace property and orthogonality of the singular vectors, $\<\mM^k, \Delta> =\<\mU \mSigma \mV^\top, u_1 v_1^\top> = \tr \diag({\sigma_1, 0, \dots, 0}) = \norms{\mM^k}$, which completes the proof.
\end{proof}

We name the derived algorithm \emph{Neon}. In the~\cref{sec:matrix-side}, we will discuss how to compute the Neon's update.

@@

Neon's and Muon's updates seem to be complete opposites: one has rank one, while the other is full-rank. It would be interesting to derive algorithms with updates of various ranks.

\textbf{Schatten norms}
\citet{cesista2025schattenp} considered Schatten-$p$ norms $\norm{\mM^k}_{S_p} = \left(\sum_{i=1}^{\min(m, n)}\sigma_i^p\right)^{1/p}$. The update was as follows: $$X^{k+1} = X^k - \eta_k \mU \frac{\diag\left(\sigma_1^{q-1},\dots, \sigma_{\min(m,n)}^{q-1}\right)}{\norm{\mM^k}^{q-1}_{S_q}}\mV^\top$$ for $q$: $\frac{1}{p} + \frac{1}{q} = 1$. This formula recovers Neon when $p\rightarrow1$ provided that $\sigma_1 > \sigma_2$, which is true on real data; NSGD when $p=2$, and Muon when $p\rightarrow\infty$.

However, Schatten norms do not fill the gap rank: indeed, when $p > 1$, the rank of the update is full, while when $p=1$ ($p \rightarrow 1$) it is one. Moreover, it is not clear how to calculate the update for $p \neq 1, 2, \infty$: it seems one has to know all $\sigma_i$ to compute the update, which makes the problem as hard as the full SVD.

\textbf{Ky Fan norms}
There is another family of matrix norms, which might help us: Ky Fan norms. For $k \in \{1, \dots, \min(m, n)\}$, the Ky Fan $k$-norm $\normkfk{\cdot}$ is $\sum_{i=1}^{k}\sigma_i$, i.e. the sum of the $k$ largest singular values of the matrix. Special cases of the Ky Fan $k$-norm are the Ky Fan $1$-norm, which is the sprectral norm, and the Ky Fan $\min\{m, n\}$-norm, which is the nuclear norm.

\textbf{Duals to Ky Fan norms}
While Schatten norms are closed under dualization and $\norm{\mM^k}_{S_p}^\dagger = \norm{\mM^k}_{S_q}$, for Ky Fan norms it is not generally the case: $\normkfk{\cdot}^\dagger = \max\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\}$ (see \citet{bhatia2013matrix}, p. 96). $k = 1$ and $k=\min(m, n)$ are exceptional: for $k=1$ the dual norm is $\normn{\cdot}$ with $k = \min(m, n)$ and for $k = \min(m, n)$ the dual norm is $\norms{\cdot}$ with $k=1$. For each other $k$ $\normkfk{\cdot} \neq \norm{\cdot}_{\text{KF-}k'}$ for any $k'$: $k'=\min(m, n)$ and $k'= 1$ correspond to the already discussed cases, while for other $k'$ one can change $\norm{\mM^k}_{\text{KF-}k'}$ by moving a small value between $\sigma_{k'}$ and $\sigma_{k'+1}$. During this change, $\norms{\mM^k}$ and $\normn{\mM^k}$ will remain constant, hence $\normkfk{\mM^k}$ will not change as well.

@@

Then prove a lemma for any set of norms \#1, \dots, \#n (by induction)

@@

\subsection{Add SignSGD: S-Muon and S-Neon}
Trivial formulas, but we need more experiments.

@@
